``{r}
#| eval: false
# Load the packages and data
library(tidyverse)
library(here)
library(janitor)
library(tidyverse)
library(tmap)
library(sf)
library(tigris)
library(leaflet)

#set wd as the 2026-hw5-team-woody folder 
sdoh <- read_csv("data/AIDSVu_County_SDOH_2023-20250726.csv")
prEP <- read_csv("data/AIDSVu_County_PrEP_2023_20250501.csv")
newdx <- read_csv("data/AIDSVu_County_NewDX_2023-20250726.csv")

```   

```{r}
#Make a value NA for data is noted as being suppressed/missing due to too low of a population size, data not shared due to a data agreement, etc

sdoh <- sdoh %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

prEP <- prEP %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

newdx <- newdx %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

```




```{r}
# converting variable from camelCase into snake_case 
sdoh  <- sdoh  %>% clean_names()
prEP <- prEP %>% clean_names()
newdx <- newdx %>% clean_names()
```



```{r}
# Use rename_with function to fix prep split in "pr_ep" to prep
prEP <- prEP %>%
  clean_names() %>%
  rename_with(
    ~ str_replace_all(.x, "pr_ep", "prep")
  )
```




```{r}
# Rate stability is listed with a Y when they are reliable rates (i.e. those generated with a numerator of 12 or greater) and N when they are not reliable. Also rate stability is listed as -9 when rate is unavailable, suppressed, or zero.  Make NAs for rates if they are any of those scenarios 

# use the variable of interest to convert N to NA
prEP <- prEP %>%
  mutate(county_prep_rate = ifelse(county_prep_rate_stability == "N", NA, county_prep_rate))

newdx <- newdx %>%
  mutate(new_diagnoses_rate   = ifelse(new_diagnoses_rate_stability == "N", NA, new_diagnoses_rate),
    new_diagnoses_black_rate   = ifelse(new_diagnoses_black_rate_stability == "N", NA, new_diagnoses_black_rate),
    new_diagnoses_white_rate   = ifelse(new_diagnoses_white_rate_stability == "N", NA, new_diagnoses_white_rate),
    new_diagnoses_hispanic_rate = ifelse(new_diagnoses_hispanic_rate_stability == "N", NA, new_diagnoses_hispanic_rate),
    new_diagnoses_asian_rate   = ifelse(new_diagnoses_asian_rate_stability == "N", NA, new_diagnoses_asian_rate)
  )
```




```{r}
# Converting geo_id and year from numeric tpo character
newdx <- newdx %>% mutate(geo_id = as.character(geo_id),
                          year = as.character(year))
prEP <- prEP %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
sdoh <- sdoh %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
```



```{r}
# Filter  each dataset to California
newdx_ca <- newdx %>%
  filter(state == "California") %>%
  rename(geoid = geo_id) %>%
  select(-contains("heterosexual"),
         -contains ("idu"),
         -contains ("transmission"),
         -contains ("msm"))

# rename variables and delete unnecessary variables
sdoh_ca <- sdoh %>%
  filter(state == "California") %>%
  rename(geoid = geo_id, county_name = county, county_urbanicity = x2023_county_urbanicity)%>%
  select(-urbanicity_year, -sdoh_year, -unemployment_year)

# rename variable county
prep_ca <- prEP %>%
  filter(state == "California") %>%
  rename(geoid = geo_id, county_name = county) 

```




```{r}
# Join the 3 dataset by keys

merged_data <- left_join(newdx_ca, prep_ca, by = c("year", "geoid", "state", "state_abbreviation", "county_name"))

hiv_california <- left_join(merged_data, sdoh_ca, by = c("year", "geoid", "state", "state_abbreviation", "county_name"))

```

 
 
 
## Visualization 1: [Title Here]


_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## Visualization 2: [Title Here]

_Describe what this plot shows, and why it's important._



```{r}
# download California shape files
# use "clean_names" to transform variables name in uppercase letter to lowercase
# rename some variable to match common in the HIV_California variable


counties_ca <- counties(state = "CA", cb = TRUE, year = 2023) %>%
  janitor::clean_names() %>%
  rename(
    city_name          = name,
    county_name        = namelsad,
    state_abbreviation = stusps,
    state              = state_name
  ) %>%
  mutate(
    geoid = as.character(geoid),
    geoid = str_pad(geoid, width = 5, pad = "0")
  )

hiv_california <- hiv_california %>%
  mutate(
    geoid = as.character(geoid),
    geoid = str_pad(geoid, width = 5, pad = "0")
  )
#Join the HIV_California data to california shape files

hiv_california_sf <- counties_ca  %>%
  left_join(hiv_california, 
            by = c("geoid", "state", "state_abbreviation", "county_name"))

# save as csv to import into the 2026-hw-5-team-woody folder

write.csv(hiv_california_sf,file = "data/hiv_california_sf.csv")
