---
title: "Final Project: [YOUR TITLE HERE]"
author: "Bao Duong and Jean Jacques Brou"
date: "2026-06-25"
params:
  number: 5
  purpose: "Final project for PUBH 6199: Visualizing Data with R"
format: 
  html:
    toc: true
    toc-location: right
    code-fold: true
    theme: flatly
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
output-dir: docs
---
```{r}
# Load libraries and data
library(dplyr)
library(tidyverse)
library(ggtext)
````

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = "figs/",
    fig.width = 7.252,
    fig.height = 4,
    comment = "#>",
    fig.retina = 3
)
```

## Introduction

_Briefly introduce your topic and research question._

## Data

# Source of the data: We will use AIDSVu Dataset 2024

# How you will access the data: https://aidsvu.org/resources/#/datasets?years=2024&locations=county

# We will focus on California. We will merge the three dataset: SODH, new diagnosis, PreP Use; then merge by county/GEOID. We will select variables of interest and remove all missing values including minus numbers before the analysis.

# Rates of PrEP use were calculated per 100,000 population 
# Import dataset 
```{r}
sdoh <- read.csv("data/AIDSVu_County_SDOH_2023-20250726.csv")
prEP <- read.csv("data/AIDSVu_County_PrEP_2023_20250501.csv")
newdx <- read.csv("data/AIDSVu_County_NewDX_2023-20250726.csv")
```  

```{r}
#Make a value NA for data is noted as being suppressed/missing due to too low of a population size, data not shared due to a data agreement, etc

sdoh <- sdoh %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))
  
prEP <- prEP %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

newdx <- newdx %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

  
# Rate stability is listed with a Y when they are reliable rates (i.e. those generated with a numerator of 12 or greater) and N when they are not reliable. Also rate stability is listed as -9 when rate is unavailable, suppressed, or zero.  Make NAs for rates if they are any of those scenarios 

prEP <- prEP %>%
  mutate(County.PrEP.Rate = ifelse(County.PrEP.Rate.Stability == "N", NA, County.PrEP.Rate))

newdx <- newdx %>%
  mutate(New.Diagnoses.Rate   = ifelse(New.Diagnoses.Rate.Stability == "N", NA, New.Diagnoses.Rate),
    New.Diagnoses.Black.Rate   = ifelse(New.Diagnoses.Black.Rate.Stability == "N", NA, New.Diagnoses.Black.Rate),
    New.Diagnoses.White.Rate   = ifelse(New.Diagnoses.White.Rate.Stability == "N", NA, New.Diagnoses.White.Rate),
    New.Diagnoses.Hispanic.Rate = ifelse(New.Diagnoses.Hispanic.Rate.Stability == "N", NA, New.Diagnoses.Hispanic.Rate),
    New.Diagnoses.Asian.Rate   = ifelse(New.Diagnoses.Asian.Rate.Stability == "N", NA, New.Diagnoses.Asian.Rate)
  )
```  
```{r}
# Merge dataset

newdx <- newdx %>% mutate(GEO.ID = as.character(GEO.ID))
prEP <- prEP %>% mutate(GEO.ID = as.character(GEO.ID))
sdoh <- sdoh %>% mutate(GEO.ID = as.character(GEO.ID))

merged_data <- left_join(newdx, prEP, by = c("GEO.ID"))

merged_data <- left_join(merged_data, sdoh, by = c("GEO.ID"))

merged_data_edit <- merged_data %>%
  select(merged_data$GEO.ID,merged_data$GEO.ID,merged_data$State,merged_data$State.Abbreviation,merged_data$County.Name)

```  



## Visualization 1: [Title Here]

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## Visualization 2: [Title Here]

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## Visualization 3: [Title Here]

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## OR Shiny app: [Title Here]

_If you choose to create a Shiny app, describe the app's purpose and functionality. Save your code inside the shiny-app/ folder, replace the URL below with your app's URL_

[Click here to open the interactive Shiny app](https://cindyhu.shinyapps.io/two-files-app/)

## Interpretation

_Summarize your key findings. What do your visualizations reveal?_

## Limitations

_Discuss any data quality or design limitations._

## Conclusion

_Wrap up your project with a takeaway message or next steps._

## References

_Include any references to data sources, packages, or literature you used. If your visualization design is inspired by someone else's work, cite those too._

1. Reference 1
2. Reference 2
3. Reference 3

