---
title: "Final Project: [YOUR TITLE HERE]"
author: "[YOUR NAMES]"
date: "2026-06-25"
params:
  number: 5
  purpose: "Final project for PUBH 6199: Visualizing Data with R"
format: 
  html:
    toc: true
    toc-location: right
    code-fold: true
    theme: flatly
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
output-dir: docs
---


```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = "figs/",
    fig.width = 7.252,
    fig.height = 4,
    comment = "#>",
    fig.retina = 3
)
```

## Introduction

_Briefly introduce your topic and research question._

## Data

_Describe your data source(s), including where it came from, the key variables, and any pre-processing steps._

prEP
newdx

```{r}
#| eval: false
# Load the packages and data
library(tidyverse)
library(here)
library(janitor)


sdoh <- read_csv(here("data/AIDSVu_County_SDOH_2023-20250726.csv"))
prEP <- read_csv(here("data/AIDSVu_County_PrEP_2023_20250501.csv"))
newdx <- read_csv(here("data/AIDSVu_County_NewDX_2023-20250726.csv"))

```   



```{r}
#Make a value NA for data is noted as being suppressed/missing due to too low of a population size, data not shared due to a data agreement, etc

sdoh <- sdoh %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

prEP <- prEP %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

newdx <- newdx %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

```




```{r}
# converting variable from camelCase into snake_case 
sdoh  <- sdoh  %>% clean_names()
prEP <- prEP %>% clean_names()
newdx <- newdx %>% clean_names()
```



```{r}
# Use rename_with function to fix prep split in "pr_ep" to prep
prEP <- prEP %>%
  clean_names() %>%
  rename_with(
    ~ str_replace_all(.x, "pr_ep", "prep")
  )
prEP
```




```{r}
# Rate stability is listed with a Y when they are reliable rates (i.e. those generated with a numerator of 12 or greater) and N when they are not reliable. Also rate stability is listed as -9 when rate is unavailable, suppressed, or zero.  Make NAs for rates if they are any of those scenarios 

# use the variable of interest to convert N to NA
prEP <- prEP %>%
  mutate(county_prep_rate = ifelse(county_prep_rate_stability == "N", NA, county_prep_rate))

newdx <- newdx %>%
  mutate(new_diagnoses_rate   = ifelse(new_diagnoses_rate_stability == "N", NA, new_diagnoses_rate),
    new_diagnoses_black_rate   = ifelse(new_diagnoses_black_rate_stability == "N", NA, new_diagnoses_black_rate),
    new_diagnoses_white_rate   = ifelse(new_diagnoses_white_rate_stability == "N", NA, new_diagnoses_white_rate),
    new_diagnoses_hispanic_rate = ifelse(new_diagnoses_hispanic_rate_stability == "N", NA, new_diagnoses_hispanic_rate),
    new_diagnoses_asian_rate   = ifelse(new_diagnoses_asian_rate_stability == "N", NA, new_diagnoses_asian_rate)
  )
```



```{r}
# replace all 0 values with NA across the 3 datasets
sdoh <- sdoh %>%
  mutate(
    across(
      where(is.numeric),
      ~ ifelse(.x == 0, NA, .x)
    )
  )

prEP <- prEP %>%
  mutate(
    across(
      where(is.numeric),
      ~ ifelse(.x == 0, NA, .x)
    )
  )

newdx <- newdx %>%
  mutate(
    across(
      where(is.numeric),
      ~ ifelse(.x == 0, NA, .x)
    )
  )

```



```{r}
# Converting geo_id and year from numeric tpo character
newdx <- newdx %>% mutate(geo_id = as.character(geo_id),
                          year = as.character(year))
prEP <- prEP %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
sdoh <- sdoh %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
```




```{r}
# Filter Filter each dataset to California
newdx_ca <- newdx %>%
  filter(state == "California")

# rename variables and delete unnecessary variables
sdoh_ca <- sdoh %>%
  filter(state == "California") %>%
  rename(county_name = county, county_urbanicity = x2023_county_urbanicity)%>%
  select(-urbanicity_year, -sdoh_year, -unemployment_year)

# rename variable county
prep_ca <- prEP %>%
  filter(state == "California") %>%
  rename(county_name = county)

```




```{r}
# Join the 3 dataset by keys

merged_data <- left_join(newdx_ca, prep_ca, by = c("year", "geo_id", "state", "state_abbreviation", "county_name"))

hiv_california <- left_join(merged_data, sdoh_ca, by = c("year", "geo_id", "state", "state_abbreviation", "county_name"))

hiv_california
 
```

 
## Visualizatiob 1: [Title Here]

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## Visualization 2: [Title Here]

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## Visualization 3: [Title Here]

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## OR Shiny app: [Title Here]

_If you choose to create a Shiny app, describe the app's purpose and functionality. Save your code inside the shiny-app/ folder, replace the URL below with your app's URL_

[Click here to open the interactive Shiny app](https://cindyhu.shinyapps.io/two-files-app/)

## Interpretation

_Summarize your key findings. What do your visualizations reveal?_

## Limitations

_Discuss any data quality or design limitations._

## Conclusion

_Wrap up your project with a takeaway message or next steps._

## References

_Include any references to data sources, packages, or literature you used. If your visualization design is inspired by someone else's work, cite those too._

1. Reference 1
2. Reference 2
3. Reference 3

