---
title: "Final Project: Identifying California counties for targeted HIV prevention services: A County-level Analysis of HIV burden, PrEP Coverage, and socioeconomic vulnerability"
author: "Bao Duong and Jean Jacques Brou"
date: "2026-06-25"
params:
  number: 5
  purpose: "Final project for PUBH 6199: Visualizing Data with R"
format: 
  html:
    toc: true
    toc-location: right
    code-fold: true
    theme: flatly
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
output-dir: docs
---
```{r}
#| label: setup
#| include: false
#| eval: false
```

```{r}
# Load the packages and data
library(tidyverse)
library(here)
library(janitor)
library(tidyverse)
library(tmap)
library(sf)
library(tigris)
library(leaflet)
```

knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = "figs/",
    fig.width = 7.252,
    fig.height = 4,
    comment = "#>",
    fig.retina = 3
)
```
## Introduction

_Briefly introduce your topic and research question._

## Data

# Source of the data: We will use AIDSVu Dataset 2024

# How you will access the data: https://aidsvu.org/resources/#/datasets?years=2024&locations=county

```{r}
sdoh <- read_csv(here("data/AIDSVu_County_SDOH_2023-20250726.csv"))
prEP <- read_csv(here("data/AIDSVu_County_PrEP_2023_20250501.csv"))
newdx <- read_csv(here("data/AIDSVu_County_NewDX_2023-20250726.csv"))

```   

```{r}
#Make a value NA for data is noted as being suppressed/missing due to too low of a population size, data not shared due to a data agreement, etc

sdoh <- sdoh %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

prEP <- prEP %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

newdx <- newdx %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

```




```{r}
# converting variable from camelCase into snake_case 
sdoh  <- sdoh  %>% clean_names()
prEP <- prEP %>% clean_names()
newdx <- newdx %>% clean_names()
```


```{r}
# Use rename_with function to fix prep split in "pr_ep" to prep
prEP <- prEP %>%
  clean_names() %>%
  rename_with(
    ~ str_replace_all(.x, "pr_ep", "prep")
  )
```




```{r}
# Rate stability is listed with a Y when they are reliable rates (i.e. those generated with a numerator of 12 or greater) and N when they are not reliable. Also rate stability is listed as -9 when rate is unavailable, suppressed, or zero.  Make NAs for rates if they are any of those scenarios 

# use the variable of interest to convert N to NA
prEP <- prEP %>%
  mutate(county_prep_rate = ifelse(county_prep_rate_stability == "N", NA, county_prep_rate))

newdx <- newdx %>%
  mutate(new_diagnoses_rate   = ifelse(new_diagnoses_rate_stability == "N", NA, new_diagnoses_rate),
    new_diagnoses_black_rate   = ifelse(new_diagnoses_black_rate_stability == "N", NA, new_diagnoses_black_rate),
    new_diagnoses_white_rate   = ifelse(new_diagnoses_white_rate_stability == "N", NA, new_diagnoses_white_rate),
    new_diagnoses_hispanic_rate = ifelse(new_diagnoses_hispanic_rate_stability == "N", NA, new_diagnoses_hispanic_rate),
    new_diagnoses_asian_rate   = ifelse(new_diagnoses_asian_rate_stability == "N", NA, new_diagnoses_asian_rate)
  )
```




```{r}
# Converting geo_id and year from numeric tpo character
newdx <- newdx %>% mutate(geo_id = as.character(geo_id),
                          year = as.character(year))
prEP <- prEP %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
sdoh <- sdoh %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
```



```{r}
# Filter Filter each dataset to California
newdx_ca <- newdx %>%
  filter(state == "California") %>%
  rename(geoid = geo_id) %>%
  select(-contains("heterosexual"),
         -contains ("idu"),
         -contains ("transmission"),
         -contains ("msm"))

# rename variables and delete unnecessary variables
sdoh_ca <- sdoh %>%
  filter(state == "California") %>%
  rename(geoid = geo_id, county_name = county, county_urbanicity = x2023_county_urbanicity)%>%
  select(-urbanicity_year, -sdoh_year, -unemployment_year)

# rename variable county
prep_ca <- prEP %>%
  filter(state == "California") %>%
  rename(geoid = geo_id, county_name = county) 

```




```{r}
# Join the 3 dataset by keys

merged_data <- left_join(newdx_ca, prep_ca, by = c("year", "geoid", "state", "state_abbreviation", "county_name"))

hiv_california <- left_join(merged_data, sdoh_ca, by = c("year", "geoid", "state", "state_abbreviation", "county_name"))

```

 
 
 
## Visualization 1: [Title Here]


_Describe what this plot shows, and why it's important._

```{r}
# collapse hiv new_diagnoses cases into high and low groups 

hiv_california <- hiv_california %>%
  mutate(
    new_diagnoses_cases_bin = if_else(
      new_diagnoses_cases >= median(new_diagnoses_cases, na.rm = TRUE),
      "High HIV diagnoses",
      "Low HIV diagnoses"
    ),
    new_diagnoses_cases_bin = factor(
      new_diagnoses_cases_bin,
      levels = c("Low HIV diagnoses", "High HIV diagnoses")
    ),
    .after = new_diagnoses_cases # with new binned variable, put after original variable 
  )

```


```{r}
 # create a mean output for the socioeconomic factors 
    radar_data <- hiv_california %>%
  mutate(across(where(is.numeric), rescale)) %>%
  filter(county == input$county_radar) %>%
  group_by(new_diagnoses_cases_bin) %>%
  summarise(
    poverty                                  = mean(poverty, na.rm = TRUE),
    percent_less_than_high_school_education  = mean(percent_less_than_high_school_education, na.rm = TRUE),
    percent_uninsured_under_65_years_old     = mean(percent_uninsured_under_65_years_old, na.rm = TRUE),
    percent_unemployed                       = mean(percent_unemployed, na.rm = TRUE)
  ) %>% 
    mutate(across(where(is.numeric), rescale))  # normalize after summarizing

# Convert to fmsb format for package 
radar_df <- radar_data %>%
  tibble::column_to_rownames("new_diagnoses_cases_bin") %>%
  as.data.frame()

# normalize 
max_row <- rep(1, ncol(radar_df))
min_row <- rep(0, ncol(radar_df))
radar_df <- rbind(max = max_row, min = min_row, radar_df)

# Clean up column names for display
colnames(radar_df) <- c("Poverty", "< HS Education", "Uninsured", "Unemployed")


## Visualization 2: 
Describe what this plot shows, and why it's important._

```{r}
# download California shape files
# use "clean_names" to transform variables name in uppercase letter to lowercase
# rename some variable to match common in the HIV_California variable


counties_ca <- counties(state = "CA", cb = TRUE, year = 2023) %>%
  janitor::clean_names() %>%
  rename(
    city_name          = name,
    county_name        = namelsad,
    state_abbreviation = stusps,
    state              = state_name
  ) %>%
  mutate(
    geoid = as.character(geoid),
    geoid = str_pad(geoid, width = 5, pad = "0")
  )

hiv_california <- hiv_california %>%
  mutate(
    geoid = as.character(geoid),
    geoid = str_pad(geoid, width = 5, pad = "0")
  )

#Join the HIV_California data to california shape files

hiv_california_sf <- counties_ca  %>%
  left_join(hiv_california, 
            by = c("geoid", "state", "state_abbreviation", "county_name"))

```

# ── Load raw data ─────────────────────────────────────────────────────────

# write.csv(hiv_california, "hiv_california.csv", row.names = FALSE)

## Visualization 3: [Title Here]

_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## OR Shiny app: [Title Here]

_If you choose to create a Shiny app, describe the app's purpose and functionality. Save your code inside the shiny-app/ folder, replace the URL below with your app's URL_

[Click here to open the interactive Shiny app](https://cindyhu.shinyapps.io/two-files-app/)

## Interpretation

_Summarize your key findings. What do your visualizations reveal?_

## Limitations

_Discuss any data quality or design limitations._

## Conclusion

_Wrap up your project with a takeaway message or next steps._

## References

_Include any references to data sources, packages, or literature you used. If your visualization design is inspired by someone else's work, cite those too._

1. Reference 1
2. Reference 2
3. Reference 3

