---
title: "Final Project: Identifying California counties for targeted HIV prevention services: A County-level Analysis of HIV burden, PrEP Coverage, and socioeconomic vulnerability"
author: "Bao Duong and Jean Jacques Brou"
date: "2026-06-25"
params:
  number: 5
  purpose: "Final project for PUBH 6199: Visualizing Data with R"
format: 
  html:
    toc: true
    toc-location: right
    code-fold: true
    theme: flatly
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
output-dir: docs
---
```{r}
#| label: setup
#| include: false
#| eval: false
```


## Introduction

Topic : Identifying California counties for targeted HIV prevention services: A County-level Analysis of HIV Burden, PrEP Coverage, and Socioeconomic Vulnerability 

Data-driven approaches can help inform targeted HIV interventions in communities with a high burden of HIV infections and low PrEP utilization. Therefore, this project seeks to identify potential priority counties in California for targeted HIV prevention services using the 2023 AIDSVu datasets. We will explore county-level patterns of HIV burden, PrEP utilization, and socioeconomic vulnerability by answering the following research questions:   
1. Which county-level socioeconomic factors (e.g., percent living in poverty,  education attainment,  uninsured rate under 65 years old, and unemployment) are associated with higher HIV diagnosis rates in California?   
2. How do  HIV diagnosis rates vary across California counties? Do counties with higher HIV diagnosis rates also have higher PrEP utilization? 
3. Do racial  disparities in HIV diagnosis rates exist across California counties?


## Data

We use 2023 AIDSVu Data collected by state and local health departments, and de-duplicated and processed by the U.S. 
Centers for Disease Control and Prevention (CDC) to meet data quality standards for comparability and reliability. 
All 50 states, the District of Columbia (DC), and U.S. territories collect comparable confidential, 
name-based case reports of HIV infection and social determinants of health. 

We downloaded the Social Determinants of Health, County New Diagnoses, and County PrEP Data. 
We restricted our analysis to California, remove unnecessary variables and merged the three data by keys 
such: "year", "geoid", "state", "state_abbreviation", "county_name". We downloaded 2023 California shape files 
and merged it to the clean data. 


```{r}
#| eval: false
# Load the packages and data
library(here)
library(janitor)
library(tidyverse)
library(tmap)
library(sf)
library(tigris)
library(leaflet)
library(htmltools)
library(scales)
library(leaflegend)

sdoh <- read_csv(here("data/AIDSVu_County_SDOH_2023-20250726.csv"))
prEP <- read_csv(here("data/AIDSVu_County_PrEP_2023_20250501.csv"))
newdx <- read_csv(here("data/AIDSVu_County_NewDX_2023-20250726.csv"))

```   



```{r}
knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = "figs/",
    fig.width = 7.252,
    fig.height = 4,
    comment = "#>",
    fig.retina = 3
)

```




```{r}
#Make a value NA for data is noted as being suppressed/missing due to too low of a population size, data not shared due to a data agreement, etc

sdoh <- sdoh %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

prEP <- prEP %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

newdx <- newdx %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

```




```{r}
# converting variable from camelCase into snake_case 
sdoh  <- sdoh  %>% clean_names()
prEP <- prEP %>% clean_names()
newdx <- newdx %>% clean_names()
```


```{r}
# Use rename_with function to fix prep split in "pr_ep" to prep
prEP <- prEP %>%
  clean_names() %>%
  rename_with(
    ~ str_replace_all(.x, "pr_ep", "prep")
  )
```




```{r}
# Rate stability is listed with a Y when they are reliable rates (i.e. those generated with a numerator of 12 or greater) and N when they are not reliable. Also rate stability is listed as -9 when rate is unavailable, suppressed, or zero.  Make NAs for rates if they are any of those scenarios 

# use the variable of interest to convert N to NA
prEP <- prEP %>%
  mutate(county_prep_rate = ifelse(county_prep_rate_stability == "N", NA, county_prep_rate))

newdx <- newdx %>%
  mutate(new_diagnoses_rate   = ifelse(new_diagnoses_rate_stability == "N", NA, new_diagnoses_rate),
    new_diagnoses_black_rate   = ifelse(new_diagnoses_black_rate_stability == "N", NA, new_diagnoses_black_rate),
    new_diagnoses_white_rate   = ifelse(new_diagnoses_white_rate_stability == "N", NA, new_diagnoses_white_rate),
    new_diagnoses_hispanic_rate = ifelse(new_diagnoses_hispanic_rate_stability == "N", NA, new_diagnoses_hispanic_rate),
    new_diagnoses_asian_rate   = ifelse(new_diagnoses_asian_rate_stability == "N", NA, new_diagnoses_asian_rate)
  )
```




```{r}
# Converting geo_id and year from numeric tpo character
newdx <- newdx %>% mutate(geo_id = as.character(geo_id),
                          year = as.character(year))
prEP <- prEP %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
sdoh <- sdoh %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
```



```{r}
# Filter Filter each dataset to California
newdx_ca <- newdx %>%
  filter(state == "California") %>%
  rename(geoid = geo_id) %>%
  select(-contains("heterosexual"),
         -contains ("idu"),
         -contains ("transmission"),
         -contains ("msm"))

# rename variables and delete unnecessary variables
sdoh_ca <- sdoh %>%
  filter(state == "California") %>%
  rename(geoid = geo_id, county_name = county, county_urbanicity = x2023_county_urbanicity)%>%
  select(-urbanicity_year, -sdoh_year, -unemployment_year)

# rename variable county
prep_ca <- prEP %>%
  filter(state == "California") %>%
  rename(geoid = geo_id, county_name = county) 

```




```{r}
# Join the 3 dataset by keys

merged_data <- left_join(newdx_ca, prep_ca, by = c("year", "geoid", "state", "state_abbreviation", "county_name"))

hiv_california <- left_join(merged_data, sdoh_ca, by = c("year", "geoid", "state", "state_abbreviation", "county_name"))

```

 
 
 
## Visualization 1: Radar Chart of Socioeonomic Variables and Comparison to High or Low HIV Diagnosis rates 

This plot shows cards that show the average percents for each county for poverty, less than high school education, uninsured rate, and unemployment rate for 2023. Also, it shows a radar chart that normalized these soceocnomic factor from 0 to 1 across all counties. It also compares to a high and low HIV diagnosis group (seperated by median). There appears to be a relationship between having a less than High School education and higher HIV diagnosis rates for California counties. There appears to not be a strong relationship for poverty, unemployment, and uninsured rates for HIV diagnosis. 


```{r}
# collapse hiv new_diagnoses cases into high and low groups 

hiv_california <- hiv_california %>%
  mutate(
    new_diagnoses_cases_bin = if_else(
      new_diagnoses_cases >= median(new_diagnoses_cases, na.rm = TRUE),
      "High HIV diagnoses",
      "Low HIV diagnoses"
    ),
    new_diagnoses_cases_bin = factor(
      new_diagnoses_cases_bin,
      levels = c("Low HIV diagnoses", "High HIV diagnoses")
    ),
    .after = new_diagnoses_cases # with new binned variable, put after original variable 
  )

```


```{r}
 # create a mean output for the socioeconomic factors 
    radar_data <- hiv_california %>%
  mutate(across(where(is.numeric), rescale)) %>%
  filter(county == input$county_radar) %>%
  group_by(new_diagnoses_cases_bin) %>%
  summarise(
    poverty                                  = mean(poverty, na.rm = TRUE),
    percent_less_than_high_school_education  = mean(percent_less_than_high_school_education, na.rm = TRUE),
    percent_uninsured_under_65_years_old     = mean(percent_uninsured_under_65_years_old, na.rm = TRUE),
    percent_unemployed                       = mean(percent_unemployed, na.rm = TRUE)
  ) %>% 
    mutate(across(where(is.numeric), rescale))  # normalize after summarizing

# Convert to fmsb format for package 
radar_df <- radar_data %>%
  tibble::column_to_rownames("new_diagnoses_cases_bin") %>%
  as.data.frame()

# normalize 
max_row <- rep(1, ncol(radar_df))
min_row <- rep(0, ncol(radar_df))
radar_df <- rbind(max = max_row, min = min_row, radar_df)

# Clean up column names for display
colnames(radar_df) <- c("Poverty", "< HS Education", "Uninsured", "Unemployed")

# Converting geo_id and year from numeric tpo character
newdx <- newdx %>% mutate(geo_id = as.character(geo_id),
                          year = as.character(year))
prEP <- prEP %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
sdoh <- sdoh %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
```

## Visualization 2: 
Describe what this plot shows, and why it's important._

```{r}
# download California shape files
# use "clean_names" to transform variables name in uppercase letter to lowercase
# rename some variable to match common in the HIV_California variable
# standardize the geoid to 5 characters in both hiv_california and counties_ca

counties_ca <- counties(state = "CA", cb = TRUE, year = 2023) %>%
  janitor::clean_names() %>%
  rename(
    city_name          = name,
    county_name        = namelsad,
    state_abbreviation = stusps,
    state              = state_name
  ) %>%
  mutate(
    geoid = as.character(geoid),
    geoid = str_pad(geoid, width = 5, pad = "0")
  )

hiv_california <- hiv_california %>%
  mutate(
    geoid = as.character(geoid),
    geoid = str_pad(geoid, width = 5, pad = "0")
  )

#Join the HIV_California data to california shape files

hiv_california_sf <- counties_ca  %>%
  left_join(hiv_california, 
            by = c("geoid", "state", "state_abbreviation", "county_name"))

```


## Visualization 3: Bar Chart of HIV diagnosis Rates among Different racial groups 

This bar chart shows the different HIV diagnosis rates by race for each county. For most counties, Black groups experience much higher HIV diagnosis rates where it was about 3-4 times more than their White counterparts. Hispanic groups had twice as high diagnosis rates as their White Counterparts. For some counties, Multiple Races, and NH Pacific Islander 
also were at least twice as high as their White counterparts. Asian groups had less HIV diagnosis rates than their White counterparts. 

```{r}
# Join the 3 dataset by keys

merged_data <- left_join(newdx_ca, prep_ca, by = c("year", "geoid", "state", "state_abbreviation", "county_name"))

hiv_california <- left_join(merged_data, sdoh_ca, by = c("year", "geoid", "state", "state_abbreviation", "county_name"))

```
 
 
 
## Visualization 1: [Title Here]




```{r}
# your code here
```



## Visualization 2: [Title Here]

The interactive map examine geographic variation in HIV diagnosis rates by county. 
We visualize HIV new diagnoses rate per 100,000 and represented 
county PrEP use rate per 100,000 with dots. Higher new hiv diagnoses rate are concentrated in larger counties in 
the south and middle of California. Counties with higher new hiv diagnoses rate have lower PrEP use
rate compared to smaller counties with lower new hiv diagnosis rate. 
The interactive map shows that Kern county has highest new HIV diagnoses rate (29) 
compared to Imperial county with highest PrEp users rate and a lower new HIV diagnoses rate (18). 


```{r}
# your code here
```



```{r}

# use st_transform to convert my data into crs
hiv_california_sf = sf::st_transform(hiv_california_sf, 4326)

# Create point layer only for prep rate
prep_points <- hiv_california_sf %>%
  sf::st_point_on_surface()

hiv_pal <- colorBin("Reds", domain = hiv_california_sf$new_diagnoses_rate, bins = c(1, 9, 15, 20, 25, 30)
                    ) # using the sf object with hiv_california_sf


prep_pal <- colorBin("black",
  domain = hiv_california_sf$county_prep_rate, bins = c(0, 100, 200, 300, 400, 500)
 
)

leaflet(hiv_california_sf) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
 
  addPolygons(fillColor = ~hiv_pal(new_diagnoses_rate ),color = "white", weight = 1, fillOpacity = 0.9, 
        
              label = ~paste0( county_name, "<br>",
        "New Diagnosis Rate:", round(new_diagnoses_rate , 1), "<br>",
        "New HIV cases:", new_diagnoses_cases, "<br>",
        "County PrEP rate:" , round(county_prep_rate, 1), "<br>",
        "County prEP Users:" , county_prep_users ) %>% lapply(htmltools::HTML)
              
      ) %>%
  
     # Dots represent county_prep_rate only
  addCircleMarkers(
    data = prep_points,
    radius = ~sqrt(county_prep_rate) / 2,
    fillColor = ~prep_pal(county_prep_rate),
    fillOpacity = 0.9,
    stroke = FALSE,
    color = "black",
    weight = 1,
  ) %>%
# The Custom Legend for New Diagnoses Rate
  addLegend(
    pal = hiv_pal,
    values = ~new_diagnoses_rate,
    title = "New HIV Diagnoses Rate", 
    position = "bottomright",
    opacity = 1,
     labFormat = labelFormat(between = " - ")
  ) %>% 
  
  # The Custom Legend for PrEP Rate
 addLegendSize(
    values = na.omit(hiv_california_sf$county_prep_rate),
    color = "black",
    fillColor = "black",
    opacity = 0.9,
    shape = "circle",
    title = "county_prep_rate",
   
    orientation = "vertical", 
    position = "bottomleft"
  ) %>%
setView(lng = -119, 
        lat = 37, 
        zoom = 5.5
        )
```



## Visualization 3: [
New HIV Diagnosis Average rate by race/ethnicity (California 2023)]

The bar chart shows California state average of new hiv diagnoses by race/ethnicity per 100, 000.
The bar chart shows a higher average rate for black and Hispanic compared to white and other minorities. 


Counties bar chart also shows the different HIV diagnosis rates by race for each county. For most counties, Black groups experience much higher HIV diagnosis rates where it was about 3-4 times more than their White counterparts. Hispanic groups had twice as high diagnosis rates as their White Counterparts. For some counties, Multiple Races, and NH Pacific Islander  also were at least twice as high as their White counterparts. Asian groups had less HIV diagnosis rates than their White counterparts.

Click the the shiny dashboard to view counties specific average. 

```{r}

hiv_race_long <- hiv_california %>%
  summarise(
    new_diagnoses_black_rate = mean(new_diagnoses_black_rate, na.rm = TRUE),
    new_diagnoses_white_rate = mean(new_diagnoses_white_rate, na.rm = TRUE),
    new_diagnoses_hispanic_rate = mean(new_diagnoses_hispanic_rate, na.rm = TRUE),
    new_diagnoses_asian_rate = mean(new_diagnoses_asian_rate, na.rm = TRUE),
    new_diagnoses_native_hawaiian_other_pacific_islander_rate =
      mean(new_diagnoses_native_hawaiian_other_pacific_islander_rate, na.rm = TRUE),
    new_diagnoses_multiple_races_rate =
      mean(new_diagnoses_multiple_races_rate, na.rm = TRUE),
    new_diagnoses_american_indian_alaska_native_rate =
      mean(new_diagnoses_american_indian_alaska_native_rate, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "race",
    values_to = "state_rate"
  ) %>%
  mutate(
    race = recode(
      race,
      new_diagnoses_black_rate = "Black",
      new_diagnoses_white_rate = "White",
      new_diagnoses_hispanic_rate = "Hispanic/Latino",
      new_diagnoses_asian_rate = "Asian",
      new_diagnoses_native_hawaiian_other_pacific_islander_rate =
        "NH / Pacific Isl.", 
      new_diagnoses_multiple_races_rate = "Multiple races",
      new_diagnoses_american_indian_alaska_native_rate =
        "Amer. Indian / AN"
    )
  )
ggplot(hiv_race_long, aes(x = race, y = state_rate, fill = race)) +
  geom_col() +
  geom_text(aes(label = round(state_rate)), vjust = 1.5, color = "white") +
  labs(
    title = "New HIV diagnosis average rate by race/ethnicity (California 2023)",
    x = "Race / ethnicity",
    y = "Diagnosis rate (per 100,000)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```


## OR Shiny app: [HIV Diagnosis Analysis Dashboard — California]


## OR Shiny app: Identifying California counties for targeted HIV prevention services: A County-level Analysis of HIV burden, PrEP Coverage, and socioeconomic vulnerability


_If you choose to create a Shiny app, describe the app's purpose and functionality. Save your code inside the shiny-app/ folder, replace the URL below with your app's URL_

[Click here to open the interactive Shiny app](https://cindyhu.shinyapps.io/two-files-app/)


## Limitations

We had to use 2023 as that was the most recent year available for AIDSVu for their HIV diagnosis, PrEP, and socioeconomic factors. A good amount of the data was suppressed due to cases being too low to share in a dataset or the state department requested to not share to AIDSVu. The dataset also shared rates for HIV diagnosis and PrEP, but the rates had a related rate stability variable that directed us to view the rate with caution due to cases being too low (for example, American Indian for many counties) and so we mutated the dataset to remove those low confidence values from the visualizations. We also did not have PrEP rate by race and so we were unable to make a stronger analysis to consider potential interplay of race to PrEP to HIV diagonosis.  Finally, our chloropleth map analyzed by counties, but perhaps there was a better way to analyze rates and relationships more meaningfully  by population hubs in California.    

## Conclusion

This analysis identified important county-level patterns in HIV burden, PrEP utilization, and socioeconomic vulnerability across California.  Counties with higher HIV diagnosis rates generally had lower high school educational attainment and substantial racial disparities, with Black and Hispanic populations experiencing disproportionately higher HIV diagnosis rates than White populations. 
Higher new HIV diagnoses rates are concentrated in larger counties in the south and middle of California.  PrEP utilization rates vary across the state; counties with higher new HIV diagnosis rates have lower PrEP use rates compared to smaller counties.
These findings can help public health agencies prioritize resources and target HIV prevention strategies, including increasing PrEP access and addressing underlying social determinants of health in the communities most affected by HIV.  Future research should examine more recent data and incorporate race-specific PrEP utilization measures to understand better and address these disparities.


## References

_Include any references to data sources, packages, or literature you used. If your visualization design is inspired by someone else's work, cite those too._

1. Aidsvsu County dataset for 2023 for new diagosis rates, PrEP use, and SDOH factors 
2. U.S. Census TIGER/Line county shapefiles
3. Radar Chart reference https://communityconnector.mathematica.org/
4. Shinyapp assistant with prompt "Give skeleton example for these three visualizations Sketch 1: Radar chart with 4 normalized socioeconomic variables and grouped by high or low HIV diagnosis, determined by if it is higher or lower than the median 
 
Sketch 2: Geographic variation in HIV diagnosis rates by county using a choropleth maps. Shading will be done by diagnosis rate and then add size of dot to show high or low PrEP use. 

Sketch 3: Grouped bar chart showing average HIV diagnosis rates among different racial groups using the dataset of all the California counties. 





