---
title: "Final Project: [YOUR TITLE HERE]"
author: "[YOUR NAMES]"
date: "2026-06-25"
params:
  number: 5
  purpose: "Final project for PUBH 6199: Visualizing Data with R"
format: 
  html:
    toc: true
    toc-location: right
    code-fold: true
    theme: flatly
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
output-dir: docs
---


```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = "figs/",
    fig.width = 7.252,
    fig.height = 4,
    comment = "#>",
    fig.retina = 3
)
```

## Introduction

_Briefly introduce your topic and research question._

## Data

_Describe your data source(s), including where it came from, the key variables, and any pre-processing steps._

prEP
newdx

```{r}
#| eval: false
# Load the packages and data
library(tidyverse)
library(here)
library(janitor)
library(tidyverse)
library(tmap)
library(sf)
library(tigris)
library(leaflet)
library(htmltools)
library(scales)
library(dplyr)
library(tidyr)
library(ggplot2)


sdoh <- read_csv(here("data/AIDSVu_County_SDOH_2023-20250726.csv"))
prEP <- read_csv(here("data/AIDSVu_County_PrEP_2023_20250501.csv"))
newdx <- read_csv(here("data/AIDSVu_County_NewDX_2023-20250726.csv"))

```   



```{r}
#Make a value NA for data is noted as being suppressed/missing due to too low of a population size, data not shared due to a data agreement, etc

sdoh <- sdoh %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

prEP <- prEP %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

newdx <- newdx %>%
  mutate(across(where(is.numeric), ~ ifelse(.x %in% -9:-1, NA, .x))) %>%
  mutate(across(where(is.character), ~ ifelse(.x %in% as.character(-9:-1), NA, .x)))

```




```{r}
# converting variable from camelCase into snake_case 
sdoh  <- sdoh  %>% clean_names()
prEP <- prEP %>% clean_names()
newdx <- newdx %>% clean_names()
```



```{r}
# Use rename_with function to fix prep split in "pr_ep" to prep
prEP <- prEP %>%
  clean_names() %>%
  rename_with(
    ~ str_replace_all(.x, "pr_ep", "prep")
  )
```




```{r}
# Rate stability is listed with a Y when they are reliable rates (i.e. those generated with a numerator of 12 or greater) and N when they are not reliable. Also rate stability is listed as -9 when rate is unavailable, suppressed, or zero.  Make NAs for rates if they are any of those scenarios 

# use the variable of interest to convert N to NA
prEP <- prEP %>%
  mutate(county_prep_rate = ifelse(county_prep_rate_stability == "N", NA, county_prep_rate))

newdx <- newdx %>%
  mutate(new_diagnoses_rate   = ifelse(new_diagnoses_rate_stability == "N", NA, new_diagnoses_rate),
    new_diagnoses_black_rate   = ifelse(new_diagnoses_black_rate_stability == "N", NA, new_diagnoses_black_rate),
    new_diagnoses_white_rate   = ifelse(new_diagnoses_white_rate_stability == "N", NA, new_diagnoses_white_rate),
    new_diagnoses_hispanic_rate = ifelse(new_diagnoses_hispanic_rate_stability == "N", NA, new_diagnoses_hispanic_rate),
    new_diagnoses_asian_rate   = ifelse(new_diagnoses_asian_rate_stability == "N", NA, new_diagnoses_asian_rate)
  )
```




```{r}
# Converting geo_id and year from numeric tpo character
newdx <- newdx %>% mutate(geo_id = as.character(geo_id),
                          year = as.character(year))
prEP <- prEP %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
sdoh <- sdoh %>% mutate(geo_id = as.character(geo_id),
                        year = as.character(year))
```



```{r}
# Filter Filter each dataset to California
newdx_ca <- newdx %>%
  filter(state == "California") %>%
  rename(geoid = geo_id) %>%
  select(-contains("heterosexual"),
         -contains ("idu"),
         -contains ("transmission"),
         -contains ("msm"))

# rename variables and delete unnecessary variables
sdoh_ca <- sdoh %>%
  filter(state == "California") %>%
  rename(geoid = geo_id, county_name = county, county_urbanicity = x2023_county_urbanicity)%>%
  select(-urbanicity_year, -sdoh_year, -unemployment_year)

# rename variable county
prep_ca <- prEP %>%
  filter(state == "California") %>%
  rename(geoid = geo_id, county_name = county) 

```




```{r}
# Join the 3 dataset by keys

merged_data <- left_join(newdx_ca, prep_ca, by = c("year", "geoid", "state", "state_abbreviation", "county_name"))

hiv_california <- left_join(merged_data, sdoh_ca, by = c("year", "geoid", "state", "state_abbreviation", "county_name"))

hiv_california

```

 
 
 
## Visualizatiob 1: [Title Here]


_Describe what this plot shows, and why it's important._

```{r}
# your code here
```

## Visualization 2: [Title Here]

_Describe what this plot shows, and why it's important._



```{r}
# download California shape files
# use "clean_names" to transform variables name in uppercase letter to lowercase
# rename some variable to match common in the HIV_California variable
# standardize the geoid to 5 characters in both hiv_california and counties_ca

counties_ca <- counties(state = "CA", cb = TRUE, year = 2023) %>%
  janitor::clean_names() %>%
  rename(
    city_name          = name,
    county_name        = namelsad,
    state_abbreviation = stusps,
    state              = state_name
  ) %>%
  mutate(
    geoid = as.character(geoid),
    geoid = str_pad(geoid, width = 5, pad = "0")
  )

hiv_california <- hiv_california %>%
  mutate(
    geoid = as.character(geoid),
    geoid = str_pad(geoid, width = 5, pad = "0")
  )
#Join the HIV_California data to california shape files

hiv_california_sf <- counties_ca  %>%
  left_join(hiv_california, 
            by = c("geoid", "state", "state_abbreviation", "county_name"))

# Visualize HIV new diagnoses rate 

```



## Visualization 3: [Title Here]

_Describe what this plot shows, and why it's important._

  
```{r}
tmap_mode("plot")
tm_shape(hiv_california_sf) +
  tm_polygons("county_prep_rate", palette = "Reds") +
                tm_layout(title = "County prep rate")
```



```{r}
tmap_mode("plot")
tm_shape(hiv_california_sf) +
  tm_polygons("new_diagnoses_rate", palette = "Reds") +
                tm_layout(title = "new diagnoses rate")
```



```{r}

# use st_transform to convert my data into crs
hiv_california_sf = sf::st_transform(hiv_california_sf, 4326)

# Create point layer only for prep rate
prep_points <- hiv_california_sf %>%
  sf::st_point_on_surface()

hiv_pal <- colorBin("Reds", domain = hiv_california_sf$new_diagnoses_rate, 
                    bins = c(0, 5, 10, 15, 20, 25, 30),
                    na.color = "gray") # using the sf object with hiv_california_sf


prep_pal <- colorBin(
  palette = "black",
  domain = hiv_california_sf$county_prep_rate,
  bins = c(0, 99, 199, 299, 399, 500),
  na.color = "gray")


leaflet(hiv_california_sf) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
 
  addPolygons(fillColor = ~hiv_pal(new_diagnoses_rate ),color = "white", weight = 1, fillOpacity = 0.9,  
        
              label = ~paste0( county_name, "<br>",
        "New Diagnosis Rate:", round(new_diagnoses_rate , 1), "<br>",
        "New HIV cases:", new_diagnoses_cases, "<br>",
        "County PrEP rate:" , round(county_prep_rate, 1), "<br>",
        "County prEP Users:" , county_prep_users ) %>% lapply(htmltools::HTML)
              
      ) %>%
  
     # Dots represent county_prep_rate only
  addCircleMarkers(
    data = prep_points,
    radius = ~sqrt(county_prep_rate) / 3,
    fillColor = ~prep_pal(county_prep_rate),
    fillOpacity = 0.9,
    stroke = TRUE,
    color = "black",
    weight = 1,
  ) %>%
addLegend(pal = hiv_pal, values = ~ new_diagnoses_rate , title = "New Diagnosis Rate", position =  "bottomright") %>% #added legend
  
  addLegend(
    pal = prep_pal,
    values = ~county_prep_rate,
    title = "County PrEP Rate",
    position = "bottomleft"
  ) %>%

setView(lng = -119, 
        lat = 37, 
        zoom = 5.5
        )
```


line break

```{r}

hiv_race_long <- hiv_california %>%
  summarise(
    new_diagnoses_black_rate = mean(new_diagnoses_black_rate, na.rm = TRUE),
    new_diagnoses_white_rate = mean(new_diagnoses_white_rate, na.rm = TRUE),
    new_diagnoses_hispanic_rate = mean(new_diagnoses_hispanic_rate, na.rm = TRUE),
    new_diagnoses_asian_rate = mean(new_diagnoses_asian_rate, na.rm = TRUE),
    new_diagnoses_native_hawaiian_other_pacific_islander_rate =
      mean(new_diagnoses_native_hawaiian_other_pacific_islander_rate, na.rm = TRUE),
    new_diagnoses_multiple_races_rate =
      mean(new_diagnoses_multiple_races_rate, na.rm = TRUE),
    new_diagnoses_american_indian_alaska_native_rate =
      mean(new_diagnoses_american_indian_alaska_native_rate, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "race",
    values_to = "rate"
  ) %>%
  mutate(
    race = recode(
      race,
      new_diagnoses_black_rate = "Black",
      new_diagnoses_white_rate = "White",
      new_diagnoses_hispanic_rate = "Hispanic/Latino",
      new_diagnoses_asian_rate = "Asian",
      new_diagnoses_native_hawaiian_other_pacific_islander_rate =
        "Native Hawaiian / Other Pacific Islander",
      new_diagnoses_multiple_races_rate = "Multiple races",
      new_diagnoses_american_indian_alaska_native_rate =
        "American Indian / Alaska Native"
    )
  )
ggplot(hiv_race_long, aes(x = race, y = rate, fill = race)) +
  geom_col() +
  geom_text(aes(label = round(rate)), vjust = 1.5, color = "white") +
  labs(
    title = "HIV diagnosis rate by race/ethnicity (California 2023)",
    x = "Race / ethnicity",
    y = "Diagnosis rate (per 100,000)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```



## OR Shiny app: [Title Here]


Dodged bar chart to display prep users, prep rate by race

_If you choose to create a Shiny app, describe the app's purpose and functionality. Save your code inside the shiny-app/ folder, replace the URL below with your app's URL_

[Click here to open the interactive Shiny app](https://cindyhu.shinyapps.io/two-files-app/)

## Interpretation

_Summarize your key findings. What do your visualizations reveal?_

## Limitations

_Discuss any data quality or design limitations._

## Conclusion

_Wrap up your project with a takeaway message or next steps._

## References

_Include any references to data sources, packages, or literature you used. If your visualization design is inspired by someone else's work, cite those too._

1. Reference 1
2. Reference 2
3. Reference 3

